A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-08-25 text/introduction.Rmd
edit for clarity

Recent advances in technology have greatly expanded our view into the mechanisms
of cancer at the systems biology level. Array and next-generation sequencing
technologies have made it possible to assay germline and somatic mutations,
expression profiles, DNA methylation, and copy number variations from the same
tissue samples. To take advantage of these tools several large consortia,
including the cancer genome atlas (TCGA) and pan-cancer analysis of whole
genomes (PCAWG), sequenced large numbers of tumors and collected data from
multiple assays that could be analyzed together, with the goal of integration
and increased understanding of the mechanisms of cancer.

Considerable progress has been made analyzing these data. Statistical analyses
identify hundreds of global and tissue-specific cancer driver genes
[@deesMuSiC2012; @tamboreroOncodriveCLUST2013;
@lawrenceDiscoverySaturationAnalysis2014;
@kumarStatisticallyIdentifyingTumor2015; @tokheimEvaluatingEvaluationCancer2016;
@jiangWITER2019; @zhaoModelingPositive2019] using approaches aimed at detecting
when genes are mutated at a greater rate than expected due to chance. It has
been estimated that fewer than five mutations in key oncogenes and/or tumor
suppressors would be sufficient to transform a normal cell to a cancerous state
[@vogelsteinPathCancerThree2015; @Iranzo-Network]. Somewhat paradoxically,
mutations in cancer driver genes are commonplace in somatic tissues and
correlate with age and environmental exposures [@mutations-healthy-tissues].

@@

More recently, attempts have been made to consider pan-cancer mutations and
other data as disruptions of normal pathway activity. @PCAWG-PATH carried out
near-exhaustive analysis of pathway enrichment using whole genome data. Network
approaches address the sparseness of mutations, allowing genes to be influenced
by the aggregate of mutations in the nearest network neighbors
[@hornNetSigNetworkbasedDiscovery2018; @Iranzo-Network]. These methods are
powerful especially when applied to a small number (on the order of hundreds) of
tumors. One of the challenges in the field remains that beyond a couple hundred
common driver genes and mutations, there may be thousands of moderate effect
genes that occur at such low frequency--given tissue diversity--and low sample
numbers that they are impossible to detect using positive selection theory. Some
groups have attempted to address this challenge by applying machine learning to
cancer data sets to discover groups of functionally related genes as they
interact with larger pathways and networks [@kimPancancerAnalysisSomatic2018;
@mourikis-ml-pathways; @moonlight]. These studies have proven very effective at
highlighting some potentially fundamental disease phenotypes at the pathway
level that cross cancers with different origins at the cellular and tissue
level.

>>>>

Recent advances in technology have greatly expanded our view into the mechanisms
of cancer at the systems biology level. Array and next-generation sequencing
technologies have made it possible to assay germline and somatic mutations,
expression profiles, DNA methylation, and copy number variations from the same
tissue samples. To take advantage of these tools, several large consortia,
including the cancer genome atlas (TCGA) and pan-cancer analysis of whole
genomes (PCAWG) sequenced large numbers of tumors and collected data from
multiple assays that could be analyzed together, with the goal of integration
and increased understanding of the mechanisms of cancer.

Considerable progress has been made analyzing these data. Statistical analyses
identify hundreds of global and tissue-specific cancer driver genes
[@deesMuSiC2012; @tamboreroOncodriveCLUST2013;
@lawrenceDiscoverySaturationAnalysis2014;
@kumarStatisticallyIdentifyingTumor2015; @tokheimEvaluatingEvaluationCancer2016;
@jiangWITER2019; @zhaoModelingPositive2019] using approaches aimed at detecting
when genes are mutated at a greater rate than expected due to chance. It has
been estimated that fewer than five mutations in key oncogenes and/or tumor
suppressors would be sufficient to transform a normal cell to a cancerous state
[@vogelsteinPathCancerThree2015; @Iranzo-Network]. Somewhat paradoxically,
mutations in cancer driver genes are commonplace in healthy tissue and
correlate with age and environmental exposures [@mutations-healthy-tissues].

@@

More recently, attempts have been made to consider pan-cancer mutations and
other data as disruptions of normal pathway activity. @PCAWG-PATH carried out
near-exhaustive analysis of pathway enrichment using whole genome data. Network
approaches address the sparseness of mutations, allowing genes to be influenced
by mutations in the nearest network neighbors
[@hornNetSigNetworkbasedDiscovery2018; @Iranzo-Network]. These methods are
powerful especially when applied to a small number (on the order of hundreds) of
tumors. One of the challenges in the field remains that beyond a couple hundred
common driver genes and mutations, there may be thousands of moderate effect
genes that occur at such low frequency--given tissue diversity--and low sample
numbers that they are impossible to detect using positive selection theory. Some
groups have attempted to address this challenge by applying machine learning to
cancer data sets to discover groups of functionally related genes as they
interact with larger pathways and networks [@kimPancancerAnalysisSomatic2018;
@mourikis-ml-pathways; @moonlight]. These studies have proven very effective at
highlighting fundamental disease phenotypes at the pathway
level across cancers with different origins at the cellular and tissue
level.

### 2020-08-25 text/manuscript.Rmd
edit for clarity

We investigated this dataset further using multiple correspondence analysis
(MCA)  [@JSSv025i01], and visually summarized the analysis with UMAP (**figure
2**) [@McInnes2018]. We used the resulting UMAP graph coordinates to perform
density based clustering with HDBSCAN [@hdbscan], which resulted in
identification of 10 well-defined clusters capturing about 80% of the tumor
samples. To capture the remaining samples into one of these 10 clusters we used
kNN (see **Supplementary Methods** for details on all clustering algorithms).

@@

Having defined tumors in terms of their pathway disruption profile, we sought to
understand whether different cancer types segregated into one or more
predominant classes. To our surprise, most cancer types were not heavily biased
in one class, and all well-represented cancer types had tumors in every class
(see **figure 3A** and full tumor profiles in **supplementary figure S1**, see
also interactive media from **supplementary file 1 (html)**), suggesting that,
in principle at least, our pathway-disruptions identify clusters of molecular
pathology largely independent of tissue-of-origin. As an example of one type of
cancer that does have a biased pathway profile, pancreatic adenocarcinoma (PAAD)
was predominantly found in class 8 (**figure 3A** and **supplementary figure
S1**). But even PAAD comprises tumors from the nine remaining classes, meaning
that patients suffering these tumors have potentially different underlying
molecular pathologies.

@@

Many cancers have molecular or histological subtypes defined based on gene
expression or other -omics profiles or pathology lab results. These subtypes
often have different standards of care owing to different overall drug
sensitivity (or other factors). If the histological subtypes represent true
molecular phenotypes, one predicts that histological subtypes should segregate
with our pathway-based classes, therefore providing support for the classes as
proxies for molecular pathology sub-typing. To our surprise, we find a similar
result to the previous analysis of cancer types projected onto the UMAP of
pathway disruptions. To illustrate this, we projected annotations for each of
the breast cancer subtypes, composed of Triple-negative/Basal-like, Her2
positive, normal-like, and luminal A and B subtypes onto the UMAP. These are
among the most heavily studied molecular subtypes in cancer, which each have
different prognoses and standards of care. We did not observe any exclusive
segregation by pathway for these subtype annotations (**Figure 3B** and
**supplemental file 1**). We also projected histological subtype data for the
remaining cancers (see **supplementary figure S2** for the full set of
projections); we find that the subtypes, though often biased towards one or more
classes, are almost never exclusive. We interpret these data in aggregate to
mean that our pathway disruption classes do not correspond to previously
identified molecular subtypes within the parent cancer type.

@@

Tumor staging is based on physico-pathological criteria, including tumor
diameter, which can vary greatly in importance between different tissues. Stage
is used clinically as a proxy for advancement toward a more deadly state and
metastasis. Given these assumptions, it is possible that more advanced tumors
have common pathway disruption profiles, and the UMAP which features a series of
lobe-like structures on a common backbone of tumor samples (**figure 2**) could
in principle also reflect progression through a series of stages. In support of
this hypothesis, the backbone starts with a cluster of tumors (class 1) that has
the fewest point mutations and culminates in a cluster (class 10) which has
nearly every pathway disrupted (**figure 4**). However, outside of class 10 we
don't observe an obvious trend in the overall mutation burden across the
backbone of the UMAP (**figure 1H**). Nonetheless, to test the hypothesis that
the molecular-pathway disruption clusters represent advancement through stages,
we projected staging data onto the UMAP. Similar to tissue of origin and other
categories of tumor, we did not observe any bias among the stages to specific
pathway disruption classes (**figure 3D**), suggesting that stage is not an
determining factor in pathway disruption classes.

@@

**Figure 4: Somatic mutation frequencies for tumors in each class** 
Column uses a logarithmic axis to represent mutation count. Row represents the
tumor samples. Each dot corresponds to the sample mutation count for tumors in
each class, ordered by mutation counts. The median mutation count in each
cluster is indicated by the horizontal line.

@@

@@

To discover what pathways are most important for clustering, we calculated
percent enrichment for each pathway _within cluster_ relative to all other
clusters combined (see methods) and ranked pathways for each cluster from
highest to lowest enrichment. We visualized the enrichment as a heatmap
(**figure 5B**). Using this approach, we identified 25 pathways highly enriched
(enrichment score > 0.3, see methods) in cluster 2, 9 pathways enriched in
cluster 3, 24 pathways in cluster 4, 10 pathways in cluster 5, 10 pathways in
cluster 6, 24 pathways in cluster 7, 24 pathways in cluster 8, 10 pathways in
cluster 9, and 0 pathways for clusters 1 and 10. Clusters 7, 8, and 9 had
several pathways in common. To explore the specific pathways marking each
cluster, we projected disruptions for each of the 377 pathways onto the UMAP
(**supplementary figure S4**). Clusters 3 and 5 were distinguished by metabolic
pathways including RNA and protein biosynthesis (**supplementary figure S4**).
Similarly, cluster 4 was distinguished by mutations affecting regulation of DNA
and histone methylation ("DNA methylation", "PRC2 methylates histones and DNA",
and "Nucleosome assembly"). Clusters 7-9 have in common mutations in
extracellular, intracellular, and immune-related signaling pathways (see
**figure 5C** and **supplementary figure S4**). Cluster 2 had the highest
pathway enrichment levels of the three, having mutations in hedgehog signaling,
"β-catenin degradation", "cellular response to hypoxia", "regulation of cell
cycle" and "apoptosis" among others.

Prior efforts to extract common signatures from pan-cancer datasets met with
difficulty in distinguishing tumor samples from their tissue-specific -omics
data signatures. Given our pathway-disruption based clustering, this raises
the question, are tumor phenotypes driven entirely by common driver genes, or by
“silent” tissue-specific effectors ( _i.e._ too few samples to detect above
statistical significance thresholds), or a combination of both? To answer this
question, we compared top pathway genes for each cluster relative to the TCGA
background to find differentially mutated genes. We ranked odds ratios and
selected the top ten enriched and depleted genes (pvalue < 0.01) for each
cluster (**Figure 6**; odds ratios plot). Clusters 7 and 8, which shared
multiple enrichment in signaling pathways, are largely driven by mutations in
_PI3K_ and its orthologs and _Ras_ genes, respectively (compare _PIK3CA_ and
_KRAS_ panels of **Supplementary figure S5**). Interestingly, cluster 9, which
also shared multiple enrichment in signaling pathways with clusters 7 and 8, is
enriched for both _PIK3CA_ and _KRAS_. This result suggests that tumors in this
cluster could be derivative of either cluster 7 or 8. Clusters 3 and 5, defined
by enrichment in metabolic pathways, had mutations in ribosomal proteins and
nuclear pore complex, respectively. Cluster 4 had mutations in genes responsible
for nucleosome structure. Cluster 2 had mutations in proteasomal subunit genes
involved in protein degradation. We also observed that genes that were enriched
for one cluster are depleted from others ( _e.g. TP53_ is enriched in cluster
6, but depleted in cluster 7; _PIK3CA_ is enriched in cluster 7, but depleted in
clusters 3 and 8). Next, we investigated the proportion of samples per cancer
type for the significant genes within a cluster (**Figure 6**; heatmap).
Surprisingly, cluster-specific tumors were not predominated by one or more
highly mutated genes across all cancers. Instead, when observing the mutation
rate for these genes within samples that belong to a cluster, the mutation rate
is heterogeneous across tumors by tissue origin ( _e.g._ In cluster 4, CESC was
enriched for _H2AFX_, OV was enriched for _HIST1H2BD_, and UCEC was enriched for
_HIST1H2AC_). Even among the top most mutationally enriched genes within
clusters there is no global pattern, indicating that our clusters are not driven
by individual genes, but rather networks as a whole. Taken together, our data
identifies a framework of cancer type-specific mutations associated with
specific clusters.

@@

**Figure 6: Gene level analysis reveals tissue-specific class signatures.** 
Odds ratio plot; Column uses a logarithmic axis to represent odds ratio with a
95% confidence interval. Row represents the significant genes from each cluster.
Each cluster was compared against the background (all other clusters) to find
deferentially mutated genes. Significant genes (pvalue < 0.01) were selected and
limited to the top ten results for each clusters. Heatmap; Column represents the
cancer type. Row represents the significant genes as described above. The
heatmap shows the proportion of samples per cluster, per cancer type that is
mutated for each of the significant genes. Depleted significant genes (left).
Enriched significant genes (right).

@@

Following the same logic we used to investigate cluster-specific enrichment of
pathways (**figure 5B**) we compared metastatic _vs._ non-metastatic
tumors, as we find that metastatic tumors are distributed across all ten
clusters ( _e.g._ **figure 3E**). Using all non-metastatic tumors as background,
we found very low levels of enrichment (≤ 10%) in a handful of pathways. We
reasoned that the individual clusters might be too different to detect global
metastasis enrichment signal given the small sample size (n = 215 metastatic
tumor samples).

Therefore, we calculated cluster-specific enrichment in metastatic tumors and
found a total of 31 enriched pathways across all clusters (**table 1**). A
number of enrichments were found in multiple clusters represented by pathways
that were already shown to be enriched in a neighboring cluster. For example,
“Signaling by _PTK6_” is enriched in non-metastatic samples of cluster 8 (see
**supplementary figure S4**) but not in 7 and 9. This pathway is enriched in
metastatic tumors of clusters 7 and 9 ($p < 10^{-3}$, **table 1**). This is also
true of “Erythropoietin activates _RAS_”, which is enriched in non-metastatic
tumors of cluster 8 (**supplementary figure S4**) and also in metastatic tumors
of clusters 7 and 9. Cluster 4 metastatic samples were enriched for “Fc epsilon
receptor (FCERI) signaling”, a key neutrophil pathway, which is also specific to
clusters 2, 7 and 8 non-metastatic tumors. Thus, pathway lesions specific to
metastases tend toward complementary mutations that are otherwise specific to
non-metastatic tumors of other clusters.

>>>>

We investigated this dataset further using multiple correspondence analysis
(MCA)  [@JSSv025i01], and visually summarized the analysis with UMAP (**figure
2**) [@McInnes2018]. We used the resulting UMAP graph coordinates to perform
density based clustering with HDBSCAN [@hdbscan], which resulted in
identification of 10 well-defined clusters capturing about 80% of the tumor
samples. To capture the remaining samples into one of these 10 clusters we used
kNN (see **Supplementary Methods** for details on clustering methods).

@@

Having defined tumors in terms of their pathway disruption profile, we sought to
understand whether different cancer types segregated into one or more
predominant classes. To our surprise, most cancer types were not heavily biased
in one class, and all well-represented cancer types had tumors in every class
(see **figure 3A** and full tumor profiles in **supplementary figure S1**, see
also interactive media from **supplementary file 1**), suggesting that,
in principle at least, our pathway-disruptions identify clusters of molecular
pathology largely independent of tissue-of-origin. As an example of one type of
cancer that does have a biased pathway profile, pancreatic adenocarcinoma (PAAD)
was predominantly found in class 8 (**figure 3A** and **supplementary figure
S1**). But even PAAD comprises tumors from the nine remaining classes, meaning
that patients suffering these tumors have potentially different underlying
molecular pathologies.

@@

Many cancers have molecular or histological subtypes defined based on gene
expression or other -omics profiles or pathology lab results. These subtypes
often have different standards of care owing to different overall drug
sensitivity (or other factors). If the histological subtypes represent true
molecular phenotypes, one predicts that histological subtypes should segregate
with our pathway-based clusters, therefore providing support for the clusters as
proxies for molecular pathology sub-typing. To our surprise, we find a similar
result to the previous analysis of cancer types projected onto the UMAP of
pathway disruptions. To illustrate this, we projected annotations for each of
the breast cancer subtypes, composed of Triple-negative/Basal-like, Her2
positive, normal-like, and luminal A and B subtypes onto the UMAP. These are
among the most heavily studied molecular subtypes in cancer, which each have
different prognoses and standards of care. We did not observe any exclusive
segregation by pathway for these subtype annotations (**Figure 3B** and
**supplemental file 1**). We also projected histological subtype data for the
remaining cancers (see **supplementary figure S2** for the full set of
projections); we find that the subtypes, though often biased towards one or more
classes, are almost never exclusive. We interpret these data in aggregate to
mean that our pathway disruption classes do not correspond to previously
identified molecular subtypes within the parent cancer type.

@@

Tumor staging is based on physico-pathological criteria, including tumor
diameter, which can vary greatly in importance between different tissues. Stage
is used clinically as a proxy for advancement toward a more deadly state and
metastasis. Given these assumptions, it is possible that more advanced tumors
have common pathway disruption profiles, and the UMAP which features a series of
lobe-like structures on a common backbone of tumor samples (**figure 2**) could
in principle also reflect progression through a series of stages. In support of
this hypothesis, the backbone starts with a cluster of tumors (class 1) that has
the fewest point mutations and culminates in a cluster (class 10) which has
nearly every pathway disrupted (**figure 4**). However, outside of class 10 we
don't observe an obvious trend in the overall mutation burden across the
backbone of the UMAP (**figure 1H**). Nonetheless, to test the hypothesis that
the molecular-pathway disruption clusters represent advancement through stages,
we projected staging data onto the UMAP. Similar to tissue of origin and other
categories of tumor, we did not observe any bias among the stages to specific
pathway disruption clusters (**figure 3D**), suggesting that stage is not a
contributing factor to cluster identity.

@@

**Figure 4: Somatic mutation frequencies for tumors in each class** 
Vertical axis shows log mutation count, horizontal axis is cluster identity.
Each dot represents an individual tumor sample, ranked lowest to highest by
mutation count. The median mutation count in each cluster is indicated by the
horizontal line.

@@

To discover what pathways are most important for clustering, we calculated
percent enrichment for each pathway _within cluster_ relative to all other
clusters combined (see methods) and ranked pathways for each cluster from
highest to lowest enrichment. We visualized the enrichment as a heatmap
(**figure 5A**). Using this approach, we identified 25 pathways highly enriched
(enrichment score ≥ 0.3, 95% confidence; see methods) in cluster 2, 9 pathways
enriched in cluster 3, 24 pathways in cluster 4, 10 pathways in cluster 5, 10
pathways in cluster 6, 24 pathways in cluster 7, 24 pathways in cluster 8, 10
pathways in cluster 9, and 0 pathways for clusters 1 and 10. Clusters 7, 8, and
9 had several pathways in common. To explore the specific pathways marking each
cluster, we projected disruptions for each of the 377 pathways onto the UMAP
(**supplementary figure S4**). Clusters 3 and 5 were distinguished by metabolic
pathways including RNA and protein biosynthesis (**supplementary figure S4**).
Similarly, cluster 4 was distinguished by mutations affecting regulation of DNA
and histone methylation ("DNA methylation", "PRC2 methylates histones and DNA",
and "Nucleosome assembly"). Clusters 7-9 have in common mutations in
extracellular, intracellular, and immune-related signaling pathways (see
**figure 5B** and **supplementary figure S4**). Cluster 2 had the highest
pathway enrichment levels of the three, having mutations in hedgehog signaling,
"β-catenin degradation", "cellular response to hypoxia", "regulation of cell
cycle" and "apoptosis" among others.

@@

Prior efforts to extract common signatures from pan-cancer datasets met with
difficulty in distinguishing tumor samples from their tissue-specific -omics
data signatures. Given our pathway-disruption based clustering, this raises
the question, are tumor phenotypes driven entirely by common driver genes, or by
“silent” tissue-specific effectors ( _i.e._ too few samples to detect above
statistical significance thresholds), or a combination of both? To answer this
question, we compared top pathway genes for each cluster relative to the TCGA
background to find differentially mutated genes. We ranked odds ratios and
selected the top ten enriched and depleted genes (pvalue < 0.01) for each
cluster (**Figure 6**; odds ratios plot). Clusters 7 and 8, which shared
multiple enrichment in signaling pathways, are largely driven by mutations in
_PI3K_ and its orthologs and _Ras_ genes, respectively (compare _PIK3CA_ and
_KRAS_ panels of **Supplementary figure S5**). Interestingly, cluster 9, which
also shared multiple enrichment in signaling pathways with clusters 7 and 8, is
enriched for both _PIK3CA_ and _KRAS_. Clusters 3 and 5, defined by enrichment
in metabolic pathways, had mutations in ribosomal proteins and nuclear pore
complex, respectively. Cluster 4 had mutations in genes responsible for
nucleosome structure. Cluster 2 had mutations in proteasomal subunit genes
involved in protein degradation. We also observed that genes that were enriched
for one cluster are depleted from others ( _e.g. TP53_ is enriched in cluster 6,
but depleted in cluster 7; _PIK3CA_ is enriched in cluster 7, but depleted in
clusters 3 and 8). Next, we investigated the proportion of samples per cancer
type for the significant genes within a cluster (**Figure 6**; heatmap).
Surprisingly, cluster-specific tumors were not predominated by one or more
highly mutated genes across all cancers. Instead, when observing the mutation
rate for these genes within samples that belong to a cluster, the mutation rate
is heterogeneous across tumors by tissue origin ( _e.g._ In cluster 4, CESC was
enriched for _H2AFX_, OV was enriched for _HIST1H2BD_, and UCEC was enriched for
_HIST1H2AC_). Even among the top most mutationally enriched genes within
clusters there is no global pattern, indicating that our clusters are not driven
by individual genes, but rather networks as a whole. Taken together, our data
identifies a framework of cancer type-specific mutations associated with
specific clusters.

@@

**Figure 6: Gene level analysis reveals tissue-specific class signatures.** 
Odds ratio plot; Column uses a logarithmic axis to represent odds ratio with a
95% confidence interval. Row represents the significant genes from each cluster.
Each cluster was compared against the background (all other clusters) to find
differentially mutated genes. Significant genes (pvalue < 0.01) were selected
and limited to the top ten results for each clusters. Heatmap; Column represents
the cancer type. Row represents the significant genes as described above. The
heatmap shows the proportion of samples per cluster and cancer type that are
mutated for each gene. Depleted significant genes (left). Enriched significant
genes (right).

@@

Following the same logic we used to investigate cluster-specific enrichment of
pathways we compared metastatic _vs._ non-metastatic tumors, as we find that
metastatic tumors are distributed across all ten clusters ( _e.g._ **figure
3E**). Using all non-metastatic tumors as background, we found very low levels
of enrichment (< 10%) in a handful of pathways. We reasoned that the individual
clusters might be too different to detect global metastasis enrichment signal
given the small sample size (n = 215 metastatic tumor samples).

Therefore, we calculated cluster-specific enrichment in metastatic tumors and
found a total of 31 enriched pathways (significant with enrichment score ≥ 0.3)
across all clusters (**table 1**). A number of enrichments were found in
multiple clusters represented by pathways that were already shown to be enriched
in a neighboring cluster. For example, “Signaling by _PTK6_” is enriched in
non-metastatic samples of cluster 8 (see **supplementary figure S4**) but not in
7 and 9. This pathway is enriched in metastatic tumors of clusters 7 and 9 ($p <
10^{-3}$, **table 1**). This is also true of “Erythropoietin activates _RAS_”,
which is enriched in non-metastatic tumors of cluster 8 (**supplementary figure
S4**) and also in metastatic tumors of clusters 7 and 9. Cluster 4 metastatic
samples were enriched for “Fc epsilon receptor (FCERI) signaling”, a key
neutrophil pathway, which is also specific to clusters 2, 7 and 8 non-metastatic
tumors. Thus, metastatic-enriched pathways from one cluster are often enriched in 
non-metastatic tumors of other clusters.

### 2020-08-25 text/survival.Rmd
edit for clarity

---
output:
  word_document: default
  html_document: default
---
If our molecular pathway disruption clusters represent distinct biological states
apart from tissue of origin, they may have different prognoses within cancer
types or across all cancers. These analyses are limited by confounding factors
of age and stage at diagnosis, sex, ethnicity, and tissue-specific disease
progression. To explore these ideas, we used Bayesian inference to test models
of survival using public longevity data from the CDC, accounting for age, gender
and ethnicity. Our model estimates the effect of cancer type and
cluster-specific cancer effects independently, resulting in a cancer and
cluster-specific estimate of the aging rate multiplier, k.

@@

The degree to which a given cancer accelerates aging can be determined by
the $k_{tis}$ multiplier (**Figure 8**) multiplied by the cluster modifier
(**Figure 9**). Looking at the tissue specific multiplier, the least deadly was
prostate cancer (PRAD), and the most deadly was glioblastoma (GBM) which has a
risk multiplier of between 15 and 22 relative to the background risk of death in
the population. This undoubtedly is influenced by the fact that relatively young
patients are affected by GBM and that it is extremely deadly. Among the
deadliest cancers outside of GBM were stomach (STAD), melanoma (SKCM) and
pancreatic (PAAD) cancers. Apart from these top 4 cancers which ranged from 6-22
in relative risk of age-adjusted death, the remaining cancer types ranged from
about 1 to 5 in magnitude. A rate of 1 represents patients who die relatively
close to the age predicted purely from their demographics independent of the
information that they have the given cancer.

@@

**Figure 8 Tissue specific effects** Effects are compared between cancer type by
showing the probability density for the posterior value of the $k_{tis}$
multiplier. These values reflect the general deadliness of the cancer type
across all clusters. GBM, PAAD, and SKCM stand out as having values in the 10 to
20 range with a wide degree of uncertainty. Other cancers such as Prostate
(PRAD) and Thyroid (THCA) show values down near or even below 1 suggesting that
these populations may have healthier lifestyles and more access to care. Most
cancer types fall in the range 2-7 as might be expected.

@@

Our estimates of tissue-specific cluster effects, in contrast, ranged from less
than 1 up to about 3 or 4, reflecting that some clusters are either less deadly
or more deadly than other clusters within each cancer type (**Figure 9**). A
cluster-specific rate of 1 represents typical rate for this tissue type. For
several cancers ( _e.g._ PRAD, kidney chromophobe (KICH), diffuse large B-cell
lymphoma (DBLC), thyroid cancer (THCA)) the posterior estimates are largely
indistinguishable from the prior, reflecting that either there were too few
mortalities in the data to make an estimate (as expected for PRAD and THCA) or
two few samples, period. We did not observe cluster-specific trends that held true
across cancer types, which could result from different cancers having different
standards of care for example. In support of this interpretation, we also tested
a factored model which considered both cancer type and each cluster independent of
cancer type (not shown). Though we were able to successfully fit this model, it
is a special case of the more general model where cluster and tissue can
independently affect longevity, and there is no reason to believe that cluster
specific effects would necessarily remain constant across tissue types given how
widely the patients vary across tissue types as well as the different drug and
surgical treatments typically used for each tissue.

>>>>

---
output:
  word_document: default
  html_document: default
---
If our molecular pathway disruption clusters represent biological states
distinct from tissue of origin, they may have different prognoses within cancer
types or across all cancers. These analyses are limited by confounding factors
of age and stage at diagnosis, sex, ethnicity, and tissue-specific disease
progression. To explore these ideas, we used Bayesian inference to test models
(see Methods and **Supplemental Methods** for greater detail) of survival using
public longevity data from the CDC, accounting for age, gender and ethnicity.
Our model estimates the effect of cancer type and cluster-specific cancer
effects independently, resulting in a cancer and cluster-specific estimate of
the aging rate multiplier, $k$.

@@

The degree to which a given cancer accelerates aging can be determined by
the $k_{tis}$ multiplier (**Figure 8**) multiplied by the cluster modifier $k_{cl}$
(**Figure 9**). Looking at the tissue specific multiplier, the least deadly was
prostate cancer (PRAD), and the most deadly was glioblastoma (GBM) which has a
risk multiplier of between 15 and 22 relative to the background risk of death in
the population. This undoubtedly is influenced by the fact that relatively young
patients are affected by GBM and that it is extremely deadly. Among the
deadliest cancers outside of GBM were stomach (STAD), melanoma (SKCM) and
pancreatic (PAAD) cancers. Apart from these top 4 cancers which ranged from 6-22
in relative risk of age-adjusted death, the remaining cancer types ranged from
about 1 to 5 in magnitude.

@@

**Figure 8 Tissue specific effects** Effects are compared between cancer type by
showing the probability density for the posterior value of the $k_{tis}$
multiplier. Larger values correspond to decreased age-independent life
expectancy.

@@

Our estimates of tissue-specific cluster effects, in contrast, ranged from less
than 1 up to about 3 or 4, reflecting that some clusters are either less deadly
or more deadly than other clusters within each cancer type (**Figure 9**). A
cluster-specific rate of 1 represents typical rate for this tissue type. For
several cancers ( _e.g._ PRAD, kidney chromophobe (KICH), diffuse large B-cell
lymphoma (DBLC), thyroid cancer (THCA)) the posterior estimates are largely
indistinguishable from the prior, reflecting that either there were too few
mortalities in the data to make an estimate (as expected for PRAD and THCA) or
two few samples, period. We did not observe cluster-specific trends that held
true across cancer types, which could result from different cancers having
different standards of care for example. In support of this interpretation, we
also tested a factored model which considered both cancer type and each cluster
independent of cancer type (not shown). Though we were able to successfully fit
this model, it is a special case of the more general model where cluster and
tissue independently affect longevity, and there is no reason to believe that
cluster specific effects would necessarily remain constant across tissue types
given how widely the patients vary across tissue types, drug and surgical
treatments.
